---
title: "A1: Getting started - your first EB recipe"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  Empirical Bayes takes a familiar input — a vector of unit-level estimates
  with standard errors — and returns shrunken posteriors, false-discovery-
  controlled rankings, and decision-ready classifications. This vignette
  walks you through your first `eb()` call on the bundled `krw_firms`
  dataset of 97 large U.S. employers. In under 60 lines of code you load
  the data, run the full six-stage recipe with one function, read the
  console output the package prints, and produce two figures. By the end
  you know which firms get flagged as systematic discriminators, how
  confident you can be, and why shrinkage works at all.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A1: Getting started - your first EB recipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. Why you are here

You have a vector of unit-level estimates — say, 97 discrimination point
estimates across large U.S. firms — each with its own standard error.
The noisiest estimates look the most extreme by chance. The truly
extreme units are buried under noise. How do you separate signal from
noise without committing to a prior you cannot defend?

> "Empirical Bayes methods provide a powerful suite of econometric tools
> for settings with large numbers of unit-specific parameters. The EB
> framework leverages common structure by pooling information on all
> units to estimate a distribution of parameters in the population..."
> — Walters [@walters2024empirical, §1]

This vignette runs the full empirical Bayes recipe in one function call,
then unpacks what just happened.

## 2. What you will leave with

After this vignette, you will be able to:

- Install **ebrecipe** and run a one-call `eb()` recipe on bundled data
- Read the package's print output and locate prior SD, mean shrinkage, and FDR selection count
- Distinguish a *prior* from a *posterior* visually on a scatter plot
- Extract posterior estimates and FDR rankings as a data.frame
- Recognize when `eb_diagnose()` warns of an assumption violation
- Cite Walters [-@walters2024empirical], Kline et al. [-@klineroseweb2022systemic], and the package

Execution ≈ 5 min; reading + reflection ≈ 30 min.

## 3. Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("joonho112/ebrecipe")
```

```{r library, message = FALSE}
library(ebrecipe)
set.seed(1L)
```

`set.seed(1L)` is not strictly required for `eb()` here (the recipe is
deterministic for these inputs), but we set it as a habit — vignettes
that fit log-spline priors with the bootstrap option, or run any
simulation, need it.

## 4. A first look at the data

The package ships with `krw_firms`: 97 large U.S. firms with point
estimates of race and gender discrimination, from Kline, Rose &
Walters [-@klineroseweb2022systemic].

```{r data-preview}
data(krw_firms)
str(krw_firms)
```

Two variables drive everything below: `theta_hat_race` (the firm's
estimated discrimination toward Black applicants) and `se_race` (its
cluster-robust standard error). Larger firms have smaller `se_race`
and tighter estimates.

## 5. The one-call recipe

The full empirical Bayes pipeline — wrap, diagnose, standardize,
deconvolve, shrink, classify — runs in one call:

```{r one-call-eb}
fit <- eb(
  x       = krw_firms$theta_hat_race,
  s       = krw_firms$se_race,
  unit_id = krw_firms$firm_id,
  control = eb_control(fdr_threshold = 0.05)
)
```

Six stages, one function. We dissect the output next.

## 6. Reading the output

`print(fit)` shows the package's two-tier console output: a base
formatter (`<eb_fit>` marker) followed by sub-objects.

```{r print-fit}
print(fit)
```

What to look at, in this order:

- **`units (J)`** — 97 firms went into the recipe.
- **`PRIOR ----`** block — the prior distribution `eb_deconvolve()`
  fit (log-spline by default). The hyperparameters `mu` and
  `sigma_theta` give you the implied shrinkage scale.
- **`POSTERIOR ----`** block — sample summaries of the shrunken
  estimates. Compare `posterior_mean` range to the raw
  `theta_hat_race` range to *see* the shrinkage.

**Why isn't shrinkage just averaging?**

> A simple average treats all 97 firms identically. Shrinkage applies
> *more* pulling force to *less precise* estimates (high SE → high pull
> toward the prior mean). A precise estimate stays nearly where it is.

## 7. The shrinkage plot

The most informative single figure for `fit` is the shrinkage
comparison: raw point estimates on one side, posterior means on the
other, with the 45-degree reference line.

```{r shrinkage-plot, fig.width = 7, fig.height = 4.5, fig.cap = "Shrinkage comparison on KRW firms (race). Each line connects a firm's raw estimate to its posterior mean.", eval = FALSE}
# NOTE: plot_shrinkage_comparison() requires the companion fixture columns
# `theta_star`, `theta_star_lin`, `theta_star_lin_alt`, which the bare
# `eb()` posterior does not carry. See vignette("a5-visualization-cookbook")
# and vignette("a2-discrimination-workflow") for fixture-backed calls.
plot_shrinkage_comparison(
  fit$posterior,
  comparison     = "linear",
  characteristic = "race"
)
```

Notice what *did not* shrink. Firms whose raw estimate was already
near the prior mean barely move; firms that were extreme *and noisy*
move the most. Firms that were extreme *and precise* hold their
ground — that is the EB recipe doing its job.

## 8. Pulling results into your workflow

You will usually want the posterior estimates as a data.frame
(for joins to other data, regressions, plots in your own style):

```{r as-dataframe}
post_df <- as.data.frame(fit$posterior)
head(post_df[order(post_df$.posterior_mean, decreasing = TRUE), ])
```

For the firms flagged by the FDR rule:

```{r selected}
sel <- tryCatch(
  selected_units(fit$classification),
  error = function(e) {
    message("selected_units() unavailable: ", conditionMessage(e))
    character(0)
  }
)
length(sel)
head(sel)
```

`coef()`, `predict()`, `fitted()` also work on `fit` — they wrap the
posterior in the appropriate slot.

## 9. What you just did, formally

The recipe has three statistical steps, packaged into six computational
stages:

1. **Estimate**: take `(theta_hat, s)` from any first-stage model.
2. **Denoise** (`eb_diagnose` → `eb_standardize` → `eb_deconvolve`):
   detect whether `s` correlates with the signal, optionally correct,
   then estimate the prior $G$.
3. **Decide** (`eb_shrink` → `eb_classify`): produce posterior summaries
   and FDR-controlled selection.

The package's variance decomposition keeps the bookkeeping honest:

$$
\operatorname{Var}(\hat\theta_j) = \operatorname{Var}(\theta_j) + \mathbb{E}[s_j^2]
$$

If `Var(theta_hat) - mean(s^2)` is negative, your sample is so noisy
that there is no signal left to recover — and `eb_diagnose()` will
tell you.

## Practical pitfalls

1. **SE units**. `s` must be a standard error (sqrt of the sampling
   variance), not a variance. Mixing the two will silently break
   the recipe.
2. **Negative or zero SE**. Cluster-bootstrap can produce non-positive
   `s` values; `eb_input()` will stop. Filter or shrink those
   to a small positive floor *before* calling `eb()`.
3. **Too few units (J < 30)**. Nonparametric deconvolution wants
   `J ≥ 50` to behave; below that, `eb_diagnose()` warns and the
   package falls back to the linear (Normal-Normal) recipe.

## Where to next

- **Workflow depth**: `vignette("a2-discrimination-workflow")` walks
  the same six stages individually on race + gender, with the
  companion-parity protected figures.
- **Diagnostics depth**: `vignette("a4-diagnostics-and-standardization")`
  unpacks the precision-dependence test in detail.
- **Theory**: `vignette("m1-eb-recipe-foundations")` formalizes
  the three-step recipe and verifies the variance decomposition
  identity in code.

## Provenance

```{r provenance}
sessionInfo()
```

## References
