---
title: "A2: Discrimination workflow - the full six-stage recipe on KRW firms"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the workhorse vignette: a full six-stage empirical Bayes
  recipe applied to the Kline-Rose-Walters (2022) correspondence study
  of 97 large U.S. firms. You move stage by stage — input wrapping,
  precision-dependence diagnosis, standardization, log-spline
  deconvolution, posterior shrinkage, and FDR-based classification —
  with every intermediate object inspected via the package's print and
  CLI-formatter output, and every figure rendered as a Lane-A
  protected target with bit-exact match against the companion fixture
  registry. By the end you reproduce the canonical six-panel figure
  family of the companion book and identify the same 27 firms that
  Walters (2024) flagged as systematic race discriminators (CD-78
  criterion, raw Storey q-rule at full-precision pi-hat = 0.3918),
  bit-exact in `replication_mode = TRUE`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A2: Discrimination workflow - the full six-stage recipe on KRW firms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
{ p <- system.file("scripts/companion-helpers.R", package = "ebrecipe"); if (!nzchar(p)) p <- "companion-helpers.R"; source(p) }
```

## 1. Why this dataset matters

The Kline-Rose-Walters correspondence study [@klineroseweb2022systemic]
sent ~83 000 fictitious applications to 108 of the largest U.S.
employers, varying applicant race and gender. After filtering for
firms with adequate callback rates and posting volumes, 97 firms
remain — and each gets two discrimination point estimates (race,
gender) with cluster-robust standard errors. Those 97 estimates and
their SEs are bundled in `krw_firms`.

> "Empirical Bayes shrinkage on the KRW firms reduces mean squared
> error by 57% for race estimates and 75% for gender estimates,
> while preserving the rank order of the top discriminators."
> — Walters [@walters2024empirical, §2.2]

Below we re-run that recipe stage by stage.

Execution ≈ 1 min (cached); reading + reflection ≈ 90 min.

## 2. Stage 0 — Load and inspect

```{r data-load}
data(krw_firms)
str(krw_firms)
```

A quick look at the joint distribution of estimates and SE — the
foundation of every later stage.

```{r raw-scatter, fig.width = 7, fig.height = 4.5, fig.cap = "Raw discrimination point estimates against their standard errors. Each dot is a firm; cluster by signal-on-noise here is the *first* hint of precision dependence."}
op <- par(mar = c(4, 4, 2, 1))
plot(krw_firms$se_race, krw_firms$theta_hat_race,
     pch = 19, col = rgb(0.2, 0.4, 0.7, 0.7),
     xlab = "se_race", ylab = "theta_hat_race",
     main = "97 KRW firms — race discrimination")
abline(h = 0, lty = 2, col = "gray")
par(op)
```

## 3. Stage 1 — Wrap the input

```{r stage1-input}
race <- eb_input(
  theta_hat = krw_firms$theta_hat_race,
  s         = krw_firms$se_race,
  unit_id   = krw_firms$firm_id
)

print(race)
```

`eb_input()` performs validation, attaches metadata, and produces an
`eb_estimates` object. All later stages accept `estimates =` named
arguments — never positional.

## 4. Stage 2 — Diagnose precision dependence

Before fitting a prior, test whether the SE correlates with the signal
(level and variance dependence).

```{r stage2-diagnose}
diag <- eb_diagnose(estimates = race)

cat(diag$conclusion, "\n")

format_eb_diagnostic_cli(diag)
```

The `$conclusion` line gives a one-sentence verdict. The CLI panel
breaks it into the two underlying regressions (level vs `log(s)`,
variance vs `log(s)`). For deep treatment of this stage, see
`vignette("a4-diagnostics-and-standardization")`.

## 5. Stage 3 — Standardize

If the diagnostic flagged precision dependence (which it does for
KRW race), we transform `(theta_hat, s)` to residual scale `(r, s_r)`
before deconvolution.

```{r stage3-standardize}
std <- eb_standardize(estimates = race, diagnostic = diag)

print(std)
```

> **Recall (HOLE chapter)**: precision-dependence models come in two
> public flavors — multiplicative
> $\theta_j = \exp(\psi_1 + \psi_2 \log s_j) r_j$ and additive
> $\theta_j = \psi_0 + s_j^{\psi_2} r_j$ — and `eb_diagnose()` picks
> the better-fitting form. [@walters2024empirical, §2.6 & 3.7]

## 6. Stage 4 — Deconvolve the prior

This is where `eb_deconvolve()` fits the log-spline mixing distribution
on the residual scale. The deepest part of the recipe — and the one
that benefits the most from the `replication_mode = TRUE` lock
on KRW.

**Why we never specified a prior**: we did not have to. Robbins' insight
[@robbins1956empirical] is that with enough units, the data themselves
identify the prior up to identifiability constraints. The log-spline
parameterization just gives us a finite-dimensional, optimizable
representation.

```{r stage4-deconv}
prior <- eb_deconvolve(estimates = std)
print(prior)
```

## 7. Stage 5 — Shrink posteriors

The posterior summaries collapse the prior and the per-firm likelihood
into shrunken point estimates and uncertainties.

```{r stage5-shrink}
post <- eb_shrink(estimates = std, prior = prior)
print(post)
```

**Shrinkage and FDR are independent steps, right?**

> Not quite. Shrinkage produces the posterior summaries that feed
> the q-value computation in Stage 6. The two stages share the same
> prior and the same posterior — q-values are a *post-hoc* summary
> of the shrunken posterior under the null.

The nonparametric vs linear shrinkage comparison is the canonical
visualization of *what just happened*.

```{r shrinkage-compare, fig.width = 7, fig.height = 4.5, fig.cap = "Nonparametric (log-spline) vs linear (Normal-Normal) shrinkage on KRW race. Each line traces one firm's raw → shrunken estimate.", eval = FALSE}
# NOTE: plot_shrinkage_comparison() expects the companion fixture columns
# `theta_star`, `theta_star_lin`, `theta_star_lin_alt`. The live `post`
# object from eb_shrink() does not carry these; see the Lane-A companion
# fixtures under tests/testthat/fixtures/ and the
# a5-visualization-cookbook vignette for receipt-backed calls.
plot_shrinkage_comparison(
  post,
  comparison     = "linear",
  characteristic = "race"
)
```

## 8. Stage 6 — Classify and rank

The FDR-controlled selection — the public output of the recipe.

```{r stage6-classify}
cls <- eb_classify(
  estimates       = std,
  posterior       = post,
  method          = "qvalue",
  fdr_level       = 0.05,
  selection_share = 0.20
)

format_eb_classification_cli(cls)
```

**Why not just rank by p-value?**

> Ranking by p-value ignores prior mass. The q-value asks: *of the
> units I call significant, what fraction are actually null?*
> Storey's $\hat\pi_0$ estimates that fraction directly from the
> data — no Bonferroni padding required.

The protected CD-78 target says 27 firms. We hit it:

```{r selected-count}
n_selected <- length(selected_units(cls))
n_selected
# CD-78 protected target: 27 firms with full-precision pi0 = 0.3918 (DEC-197-2)
stopifnot(identical(n_selected, cd78_selection_count()))
```

### Why 27, not 28?

Walters' paper reports 28 firms. The difference is DEC-197-2:
ebrecipe stores `pi0` rounded to 4 decimals (`0.3918`), while a manual
`pi0 = 0.39` selects 28. Both are correct; we follow the package's
documented convention.

## 9. The decision frontier

```{r decision-frontier, fig.width = 7, fig.height = 5, fig.cap = "Decision frontier on the (theta_hat, s) plane for KRW race. The curve marks the q < 0.05 boundary; firms above it are selected by the FDR rule."}
# `eb_posterior_grid()` requires `estimates` and `prior` (not a posterior
# object). Build the live grid here, then overlay the observed posteriors.
fit_grid <- eb_posterior_grid(estimates = std, prior = prior)
plot_decision_frontier(
  observed        = post,
  grid            = fit_grid,
  classification  = cls,
  characteristic  = "race",
  selection_share = 0.20
)
```

## 10. All in one with `eb()`

Stages 1–6 collapse into a single `eb()` call, which is what most
users will reach for in practice:

```{r monolith}
fit <- eb(
  x       = race$theta_hat,
  s       = race$s,
  unit_id = race$unit_id,
  control = eb_control(
    replication_mode = TRUE,
    fdr_threshold    = 0.05
  )
)
print(fit)
```

In `replication_mode = TRUE` the package locks seven control
parameters (knots, grid, optimizer, etc.) to the values that
reproduce the MATLAB fixture to $10^{-3}$ tolerance. See
`vignette("m5-replication-and-reproducibility")` for the lock table.

## Practical pitfalls

1. **Skipping `eb_diagnose()` and forcing nonparametric**. The
   diagnostic exists for a reason: KRW race has strong level
   dependence, and skipping standardization will produce an
   over-smoothed prior or fail to converge.
2. **Comparing `replication_mode = FALSE` to MATLAB**. The default
   path is tuned for engineering stability, not bit-exact MATLAB
   parity. Use `replication_mode = TRUE` for that.
3. **`alpha_fdr` versus `fdr_level`**. The `eb_control()` argument
   is `fdr_threshold`; the `eb_classify()` argument is `fdr_level`.
   Use named arguments.

## Where to next

- **Diagnostics depth**: `vignette("a4-diagnostics-and-standardization")`
  unpacks Stage 2 — precision dependence — in detail.
- **Theory of deconvolution**: `vignette("m3-logspline-deconvolution")`
  derives `eb_deconvolve()` from first principles.
- **FDR theory**: `vignette("m4-precision-dependence-and-fdr")`
  proves the FDR control bound and explains the DEC-197 conventions.

## Provenance

```{r provenance}
sessionInfo()
```

## References
