---
title: "A3: School value-added workflow - linear EB on simulated and bundled data"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  School value-added models (VAMs) are the canonical linear-EB
  application: school-level random effects shrunken toward a common
  mean using Normal-Normal closed forms. This vignette teaches the
  workflow through three increasingly realistic paths — fitting
  `eb_vam()` directly on student-level simulated data, importing a
  pre-aggregated school summary, and adjusting for a charter indicator
  via conditional EB. Because the original Boston administrative data
  is restricted-access, we use the bundled `vam_simulated` and
  `vam_schools` datasets, both calibrated to match Angrist et al.
  (2017). You will recover the true school effects and visualize how
  shrinkage corrects the over-dispersion of raw point estimates.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A3: School value-added workflow - linear EB on simulated and bundled data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
{ p <- system.file("scripts/companion-helpers.R", package = "ebrecipe"); if (!nzchar(p)) p <- "companion-helpers.R"; source(p) }
```

## 1. Why VAM is the canonical linear-EB example

When the prior $G$ on school effects is approximately Gaussian, the
posterior has a closed form: a precision-weighted convex combination
of the raw estimate and the prior mean. This is the *linear empirical
Bayes* (Normal-Normal) recipe, made famous in education by Angrist
et al. [-@angrist2017leveraging] and reviewed in Walters
[-@walters2024empirical, §2.8].

> "If you imagine value-added scores as samples from a population
> distribution of true school effects, then your best guess about
> any one school's effect is a weighted average of its observed score
> and the population mean — weighted by precision."
> — Walters [@walters2024empirical, paraphrasing §2.8]

Execution ≈ 1 min; reading + reflection ≈ 75 min.

## 2. The data: simulated stand-in for Boston

Two bundled datasets back this vignette:

```{r data-load}
data(vam_simulated)   # 2500 students × 50 schools (DGP truth known)
data(vam_schools)     # 50 school summaries (theta_hat, se, charter)

str(vam_simulated)
str(vam_schools)
```

**Disclosure**: `vam_simulated` is calibrated to the Angrist et al.
[-@angrist2017leveraging] benchmark on Boston schools, but the
original administrative data is restricted-access. The simulation
gives us ground truth (`true_school_fx`), which lets us *verify*
shrinkage rather than just assert it.

## 3. The math in 5 lines

For a Gaussian prior $\theta_j \sim \mathcal{N}(\mu_\theta, \sigma_\theta^2)$
and observation model $\hat\theta_j \mid \theta_j, s_j \sim \mathcal{N}(\theta_j, s_j^2)$,
the posterior is closed-form:

$$
\theta_j^* = \kappa_j \hat\theta_j + (1 - \kappa_j) \mu_\theta, \qquad
\kappa_j = \frac{\sigma_\theta^2}{\sigma_\theta^2 + s_j^2}
$$

The reliability $\kappa_j$ is *learned* from the data via method of
moments. For derivation and verification, see
`vignette("m2-linear-eb-normal-normal")`.

## 4. Path A — Student-level fit

The most direct entry point: hand `eb_vam()` student-level data and
a formula for the school grouping.

```{r path-a-fit}
fit_a <- eb_vam(
  y ~ x | school_id,
  data   = vam_simulated,
  method = "linear"
)
print(fit_a)
```

## 5. Reading the VAM output

The `<eb_vam_fit>` marker wraps an `<eb_fit>` for the linear EB
posterior. Key fields:

- **`method = linear`** — Normal-Normal closed form (no log-spline)
- **`PRIOR`** — `mu`, `sigma_theta_sq` (positive-part MoM)
- **`POSTERIOR`** — posterior means + `shrinkage_weight` $\kappa_j$

The CLI-decorated panel (optional, requires the `cli` package):

```{r path-a-cli, eval = requireNamespace("cli", quietly = TRUE)}
format_eb_vam_fit_cli(fit_a)
```

## 6. Path B — Imported school summary

In practice you often have only school-level point estimates and
standard errors (not the raw student data). `eb_vam()` accepts that
shape too:

```{r path-b-fit}
fit_b <- eb_vam(
  theta_hat ~ 1 | school_id,
  data       = vam_schools,
  se_source  = "vce_matrix",
  vce_matrix = diag(vam_schools$se^2),
  method     = "linear"
)
print(fit_b)
```

Path A and Path B should produce nearly identical posteriors when
the inputs match — verifying that the package's two entry points
agree.

## 7. Path C — Conditional EB (charter adjustment)

Charter schools may have a systematically different prior mean from
traditional public schools. Conditional EB lets the prior depend on
a covariate.

```{r path-c-fit}
fit_c <- eb_vam(
  theta_hat ~ 1 | school_id,
  data           = vam_schools,
  se_source      = "vce_matrix",
  vce_matrix     = diag(vam_schools$se^2),
  conditional_on = ~ charter,
  method         = "linear"
)
print(fit_c)
```

**Why is my charter school's posterior shifted more?**

> Two reasons combine: (a) charter schools have smaller samples
> (lower precision → more shrinkage), (b) the conditional prior
> for charter schools differs from non-charter. Both effects push
> the posterior away from the raw estimate.

## 8. Visualization triplet

```{r prior-posterior-plot, fig.width = 7, fig.height = 4.5, fig.cap = "Prior vs posterior on simulated Boston-calibrated VAM data. Shrinkage tightens the spread; the prior mean is recovered as the center of mass."}
plot_vam_prior_posterior(fit_a, method = "unconditional")
```

```{r truth-shrinkage-plot, fig.width = 7, fig.height = 4.5, fig.cap = "Truth vs raw vs shrunken posterior. Because this is simulated data, we have ground truth — and posterior means visibly track the true effects better than the raw estimates do."}
plot_vam_truth_shrinkage(fit_a, truth = vam_simulated)
```

```{r conditional-prior-plot, fig.width = 7, fig.height = 4.5, fig.cap = "Conditional EB: separate priors for charter vs non-charter schools. The two prior means differ; the posteriors inherit that difference."}
plot_vam_prior_posterior(fit_c, method = "conditional")
```

## 9. When to escalate to nonparametric

Linear EB assumes the prior is Gaussian. For Boston VAM that is
approximately true; for fat-tailed problems like firm-level
discrimination, it is not. If `eb_diagnose()` on your data warns
of departures from Gaussianity, escalate to the nonparametric path
(`method = "nonparametric"` or the full `eb()` workflow). See
`vignette("a2-discrimination-workflow")` for a worked example.

## Practical pitfalls

1. **Simulated ≠ real**. `vam_simulated` is calibrated, not Boston
   itself. The simulation is for *workflow verification*; external
   validity claims need real-data sensitivity analyses.
2. **Charter endogeneity**. `conditional_on = ~ charter` allows
   the prior to *shift* with charter status; it does *not* identify
   causal effects of charter status on student outcomes.
3. **Small J (< 30 schools)**. Even linear EB's MoM estimator is
   unstable when J is small. `vam_schools` has J = 50, comfortably
   in range.

## Where to next

- **Basics revisit**: `vignette("a1-getting-started")` if you have
  not seen the one-call `eb()` yet.
- **Linear EB theory**: `vignette("m2-linear-eb-normal-normal")`
  derives the closed forms and the James–Stein bridge.
- **Nonparametric escalation**: `vignette("a2-discrimination-workflow")`
  shows the full six-stage recipe.
- **Log-spline deconvolution theory**: `vignette("m3-logspline-deconvolution")`
  explains *when* a Gaussian prior is insufficient and what the
  nonparametric path replaces it with.

## Provenance

```{r provenance}
sessionInfo()
```

## References
