---
title: "A4: Diagnostics and standardization - handling precision dependence"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  Empirical Bayes' classic derivation assumes the noise s_j is
  *independent* of the parameter theta_j. In real audit, value-added,
  and policy datasets, that independence often fails — small firms
  have larger SEs *and* more extreme estimates, or vice versa. This
  vignette teaches you to detect precision dependence with
  `eb_diagnose()` and standardize the data with `eb_standardize()` so
  deconvolution remains valid. We compare the additive and
  multiplicative parametric families, fit psi-hat via NLLS through
  `precision_fit()`, and visualize corrected versus uncorrected priors.
  We also show why a small p-value alone is misleading — magnitude of
  psi-hat-2 matters — and what to do when J is too small for the
  diagnostic to be powerful. The narrative emphasizes that diagnostics
  are not optional bureaucracy: they are how you avoid a confidently
  wrong inference.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A4: Diagnostics and standardization - handling precision dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

> *Prerequisite*: this vignette assumes you have run the basic
> workflow in `vignette("a2-discrimination-workflow")` at least once.
> If you have not, start with `vignette("a1-getting-started")`.

## 1. The precision-independence assumption

The classic EB recipe assumes:

$$
\hat\theta_j \mid \theta_j, s_j \sim \mathcal{N}(\theta_j, s_j^2), \qquad
\theta_j \perp s_j
$$

That is: the *true* effect $\theta_j$ is independent of the
*sampling noise scale* $s_j$. When the assumption fails, even a
perfectly fit prior produces biased posteriors.

> "Two functional families capture the most common patterns of
> precision dependence:
>
> - **Multiplicative**: $\theta_j = \exp(\psi_1 + \psi_2 \log s_j) r_j$
> - **Additive**: $\theta_j = \psi_0 + s_j^{\psi_2} r_j$
>
> where $r_j$ is a precision-independent residual."
> — Walters [@walters2024empirical, §2.6, 3.7]

Execution ≈ 30 sec; reading + reflection ≈ 75 min.

## 2. What it looks like

Three flavors of `(theta_hat, s)` scatter, side by side:

```{r diagnostic-gallery, fig.width = 9, fig.height = 3.5, fig.cap = "Three diagnostic scatter patterns. Left: clean (no precision dependence). Middle: funnel (multiplicative). Right: tilted (additive)."}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))

set.seed(1L)

# Clean
s <- runif(80, 0.02, 0.10)
th_clean <- rnorm(80, 0, 0.05) + rnorm(80, 0, s)
plot(s, th_clean, pch = 19, col = rgb(0.2, 0.5, 0.3, 0.7),
     xlab = "s", ylab = expression(hat(theta)), main = "Clean")
abline(h = 0, lty = 2, col = "gray")

# Multiplicative (funnel)
th_mult <- exp(-0.2 + 1.5 * log(s)) * rnorm(80) + rnorm(80, 0, s)
plot(s, th_mult, pch = 19, col = rgb(0.7, 0.3, 0.2, 0.7),
     xlab = "s", ylab = expression(hat(theta)), main = "Multiplicative")
abline(h = 0, lty = 2, col = "gray")

# Additive
th_add <- 0.05 + s^0.8 * rnorm(80, 0, 0.5) + rnorm(80, 0, s)
plot(s, th_add, pch = 19, col = rgb(0.4, 0.2, 0.6, 0.7),
     xlab = "s", ylab = expression(hat(theta)), main = "Additive")
abline(h = 0, lty = 2, col = "gray")

par(op)
```

KRW race resembles the middle panel — a strong funnel. KRW gender
resembles the right panel — a weak tilt.

## 3. Running the diagnostic

```{r run-diagnose}
data(krw_firms)

race <- eb_input(
  theta_hat = krw_firms$theta_hat_race,
  s         = krw_firms$se_race,
  unit_id   = krw_firms$firm_id
)

diag <- eb_diagnose(estimates = race)
cat(diag$conclusion, "\n")
```

The one-sentence `$conclusion` is the headline. The CLI-decorated
panel below breaks it into the two regressions (level vs `log(s)`,
variance vs `log(s)`):

```{r diag-cli, eval = requireNamespace("cli", quietly = TRUE)}
format_eb_diagnostic_cli(diag)
```

## 4. Reading the recommendation

**Isn't precision dependence the same as heteroskedasticity?**

> Related but not identical. Heteroskedasticity = the *noise* variance
> changes with predictors. Precision dependence in EB = the *signal*
> $\theta$ also varies systematically with the noise scale $s$.
> Standardization removes the second; weighted regression addresses
> the first.

The diagnostic reports:

- **Conclusion**: a one-sentence verdict (level, variance, both, or none)
- **Level test**: regression of $\hat\theta$ on $\log(s)$
- **Variance test**: regression of $(\hat\theta - \mu)^2 - s^2$ on $\log(s)$

You should care about *magnitude*, not just p-values. With J = 97
firms, even $\psi_2 = 0.05$ can be "significant".

## 5. Refitting with `precision_fit()`

For finer control of the standardization model:

```{r precision-fit, eval = requireNamespace("cli", quietly = TRUE)}
pf <- precision_fit(diag)
format_eb_precision_fit_cli(pf)
```

`pf` exposes the fitted coefficients `psi_0, psi_1, psi_2`, their
standard errors, and the residual structure used downstream.

## 6. Standardizing the estimates

Once the model is chosen, `eb_standardize()` produces residual-scale
estimates `(r, s_r)`:

```{r standardize}
std <- eb_standardize(estimates = race, diagnostic = diag)
summary(std)
```

The summary should show `theta_hat` (now on residual scale) with
mean near zero and SD roughly 1 — the deconvolution target.

## 7. What happens if you skip standardization

Compare priors fit with vs without standardization. The unstandardized
prior absorbs the level dependence as spurious spread, giving
over-shrunken posteriors for small firms.

```{r diag-plot, fig.width = 7, fig.height = 4.5, fig.cap = "Diagnostic scatter for KRW race: raw theta_hat vs log(s), with the fitted multiplicative model overlay."}
plot(diag)
```

## 8. Custom starting values for hard cases

For weakly-identified data (KRW gender is the canonical example),
the NLLS fit can be sensitive to starting values. `precision_fit()`
accepts a `model =` argument to choose between `multiplicative`,
`additive`, and `none`:

```{r custom-precision-fit}
# Force additive even if diagnose recommends multiplicative
pf_add <- precision_fit(diag, model = "additive")
```

Convergence is recorded in `pf$convergence` (0 = success).

## 9. Sample-size caution (J < 50)

With J < 50, even the diagnostic regressions are underpowered.
A simulated demonstration:

```{r small-j-demo}
# Illustrative small-J case: simulate 20 schools, fit linear VAM,
# then diagnose the school-level estimates.
sim_small  <- eb_simulate(J = 20, sigma_theta = 0.10)
fit_small  <- eb_vam(y ~ x | school_id, data = sim_small$students,
                     method = "linear")
diag_small <- eb_diagnose(estimates = fit_small$estimates)
diag_small$conclusion
```

In small-J regimes, prefer the linear EB recipe (`method = "linear"`
in `eb()` or `eb_vam()`) — it has fewer degrees of freedom to overfit.

## Practical pitfalls

1. **p-value alone is misleading**. Even `p < 0.001` for `psi_2`
   does not justify standardization if $\hat\psi_2$ is tiny in
   magnitude.
2. **Standardized scale loses interpretability**. Residual-scale
   estimates are useful for deconvolution and shrinkage, but
   reporting needs the original scale — use `eb_change_of_variables()`
   to back-transform.
3. **Gender weak-identification (KRW)**. Documented in companion
   lessons-learned: gender $\psi_2$ NLLS is sensitive to starting
   values. Always inspect `pf$convergence` and `pf$psi_se`.

## Where to next

- **Apply the fix**: return to `vignette("a2-discrimination-workflow")`
  and re-run with the recommended model.
- **FDR + decision rules**: `vignette("a5-visualization-cookbook")`
  shows the decision frontier with corrected priors.
- **Theory**: `vignette("m4-precision-dependence-and-fdr")` derives
  the conditional prior $G(\cdot \mid s)$ and the back-transform
  Jacobian.

## Provenance

```{r provenance}
sessionInfo()
```

## References
