---
title: "A5: Visualization cookbook - recipes for every plot function"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  ebrecipe ships thirteen plot functions, each answering a different
  visual question — from "what does the prior look like?" to "did
  shrinkage preserve ordering?" This cookbook presents each function in
  a uniform format: one-line purpose, lead chunk, default plot, one
  customization variant. We organize by purpose (companion-parity,
  VAM-specific, composite dashboards, theming) rather than alphabet,
  so you navigate by what you need to communicate. No equations; pure
  visual literacy. Treat this as a reference you return to whenever
  you need the right figure for the right audience.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A5: Visualization cookbook - recipes for every plot function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. How to read this cookbook

Each recipe has a uniform template:

1. **Purpose** (one line)
2. **Lead chunk** (3–6 lines, ready to copy)
3. **Default plot**
4. **One customization variant**

We rely on a single shared fit throughout — KRW race — to keep
the cookbook concise. Substitute your own `fit` object freely.

```{r shared-fit}
data(krw_firms)

# Standardize via the multiplicative precision-dependence model so the
# resulting `fit$prior` carries the psi_1, psi_2, and spline_info needed
# by the decision-frontier and dashboard plots below.
fit <- eb(
  x       = krw_firms$theta_hat_race,
  s       = krw_firms$se_race,
  unit_id = krw_firms$firm_id,
  control = eb_control(
    fdr_threshold   = 0.05,
    standardize     = TRUE,
    precision_model = "multiplicative"
  )
)

# Companion theta-scale prior, used by the mixing-distribution and
# posterior-overlay recipes (Sections 2.1, 2.2). The shared evaluation
# point `s` is the mean standard error.
prior_theta <- eb_change_of_variables(
  prior = fit$prior,
  s     = mean(krw_firms$se_race),
  psi_1 = fit$prior$spline_info$psi_1,
  psi_2 = fit$prior$spline_info$psi_2,
  model = "multiplicative"
)
```

## 2. Companion-parity plots (Lane A)

Five functions reproduce the Walters [-@walters2024empirical, §3]
canonical figures, with bit-exact match against the protected target
registry.

### 2.1 `plot_mixing_distribution()`

Purpose: the estimated prior $G$ on its support.

```{r plot-mixing, fig.width = 7, fig.height = 4.5, fig.cap = "Mixing distribution g(r) on the residual scale (Lane A protected target g_r_white)."}
plot_mixing_distribution(fit$prior, characteristic = "white", scale = "r")
```

Customization: switch to the theta scale.

```{r plot-mixing-theta, fig.width = 7, fig.height = 4.5, fig.cap = "Same mixing distribution on the theta scale (Lane A protected target g_theta_white)."}
# Theta-scale plotting requires the prior on the theta scale. We
# pre-converted `prior_theta` in the shared-fit chunk above via
# `eb_change_of_variables()`.
plot_mixing_distribution(prior_theta, characteristic = "white", scale = "theta")
```

### 2.2 `plot_posterior_overlay()`

Purpose: posterior densities for each unit, overlaid with the prior.

```{r plot-posterior, fig.width = 7, fig.height = 4.5, fig.cap = "Posterior overlay — each curve is one firm's posterior; the thick line is the prior g."}
# `plot_posterior_overlay()` consumes a theta-scale density; we pass the
# pre-converted `prior_theta` defined in the shared-fit chunk.
plot_posterior_overlay(fit$posterior, density = prior_theta,
                       characteristic = "white")
```

### 2.3 `plot_shrinkage_comparison()`

Purpose: nonparametric vs linear shrinkage side by side.

```{r plot-shrink, fig.width = 7, fig.height = 4.5, fig.cap = "Each line connects a firm's raw estimate to its NP and linear posterior means. Diverging lines flag firms where the nonparametric correction matters most.", eval = FALSE}
# Not evaluated: `plot_shrinkage_comparison()` requires a posterior frame
# that carries both the NP estimate (`theta_star`) and the linear
# comparator column (`theta_star_lin`). The companion fixtures under
# `tests/testthat/fixtures/posteriors_white.csv` provide both; the live
# `eb()` posterior does not. The companion-parity test suite enforces
# bit-exact agreement against those fixtures.
plot_shrinkage_comparison(fit$posterior, characteristic = "white")
```

### 2.4 `plot_fdr_histogram()`

Purpose: p-value and q-value histograms with the Storey $\hat\pi_0$
rectangle overlay.

```{r plot-fdr, fig.width = 7, fig.height = 4.5, fig.cap = "FDR diagnostic: p-value histogram with the Storey pi-hat overlay (q-value panel is available via `metric = \"q\"`)."}
plot_fdr_histogram(
  posterior      = fit$posterior,
  classification = fit$classification,
  metric         = "p",
  characteristic = "white"
)
```

### 2.5 `plot_decision_frontier()`

Purpose: the (theta_hat, s) decision frontier with selected units.

```{r plot-frontier, fig.width = 7, fig.height = 5, fig.cap = "Decision frontier on the (theta_hat, s) plane. Curve = q=0.05 boundary; firms above = selected by FDR rule."}
# `plot_decision_frontier()` consumes standardization metadata from
# `fit$prior$spline_info`; the shared fit above sets
# `precision_model = "multiplicative"` so the metadata is present.
plot_decision_frontier(
  observed        = fit$posterior,
  grid            = eb_posterior_grid(fit$estimates, fit$prior),
  classification  = fit$classification,
  characteristic  = "white",
  selection_share = 0.05
)
```

## 3. VAM-specific plots (Lane B)

Two functions for the linear-EB VAM workflow:

### 3.1 `plot_vam_prior_posterior()`

```{r plot-vam-pp, fig.width = 7, fig.height = 4.5, fig.cap = "VAM prior and posterior on the same axes. Shrinkage visibly tightens the spread."}
data(vam_simulated)
fit_vam <- eb_vam(y ~ x | school_id, data = vam_simulated, method = "linear")
plot_vam_prior_posterior(fit_vam, method = "unconditional")
```

### 3.2 `plot_vam_truth_shrinkage()`

```{r plot-vam-truth, fig.width = 7, fig.height = 4.5, fig.cap = "Truth vs raw vs shrunken on simulated VAM data. Shrinkage moves estimates toward the prior, but preserves ordering of true effects."}
plot_vam_truth_shrinkage(fit_vam, truth = vam_simulated)
```

## 4. Composite dashboards

Three functions assemble multiple panels via patchwork. Useful for
papers' "headline figure" and for at-a-glance review.

### 4.1 `plot_results()`

```{r plot-results, fig.width = 9, fig.height = 6, fig.cap = "Results dashboard: prior + posterior overlay + shrinkage + selection counts in a single layout."}
plot_results(fit, characteristic = "white", scale = "r")
```

### 4.2 `plot_diagnostics()`

```{r plot-diagnostics, fig.width = 9, fig.height = 6, fig.cap = "Diagnostics dashboard: precision-dependence scatter + fitted curve + FDR check."}
plot_diagnostics(fit)
```

### 4.3 `plot_decision()`

```{r plot-decision, fig.width = 9, fig.height = 6, fig.cap = "Decision dashboard: q-value histogram + frontier + selected-unit summary."}
# The shared fit above carries `precision_model = "multiplicative"`, so
# the frontier panel inside `plot_decision()` has the metadata it needs.
plot_decision(
  observed       = fit$posterior,
  grid           = eb_posterior_grid(fit$estimates, fit$prior),
  classification = fit$classification,
  characteristic = "white"
)
```

## 5. Theme and palette

Two helpers for journal-consistent styling:

### 5.1 `theme_ebrecipe()`

```{r theme-demo, fig.width = 7, fig.height = 4.5, fig.cap = "theme_ebrecipe() applied to a ggplot. Minimal, with clear axes and the ebrecipe palette as default colors."}
library(ggplot2)
ggplot(krw_firms, aes(se_race, theta_hat_race)) +
  geom_point(color = ebrecipe_palette()[1], alpha = 0.7) +
  geom_hline(yintercept = 0, lty = 2, color = "gray") +
  labs(x = "se_race", y = "theta_hat_race", title = "KRW race — themed") +
  theme_ebrecipe()
```

### 5.2 `ebrecipe_palette()`

```{r palette-demo, fig.width = 7, fig.height = 2.4, fig.cap = "ebrecipe palette swatches."}
pal <- ebrecipe_palette()
op <- par(mar = c(1, 1, 2.5, 1))
plot(seq_along(pal), rep(1, length(pal)), pch = 15, cex = 6,
     col = pal, xlab = "", ylab = "", axes = FALSE,
     main = "ebrecipe palette")
par(op)
```

## 6. `autoplot.eb_fit()` — one-line summary

When you want a quick overview without choosing a recipe:

```{r autoplot-demo, fig.width = 9, fig.height = 6, fig.cap = "autoplot.eb_fit() with type = 'results' dispatches to the results dashboard."}
# `autoplot.eb_fit()` accepts an explicit type plus the canonical
# `characteristic` value used by the companion fixtures.
ggplot2::autoplot(fit, type = "results", characteristic = "white")
```

## 7. Customization patterns

### 7.1 Title and subtitle

```{r custom-titles, fig.width = 7, fig.height = 4.5, fig.cap = "Customized titles via standard ggplot2 +labs().", eval = FALSE}
# Not evaluated: relies on `plot_shrinkage_comparison()`, which
# requires the `theta_star_lin` comparator column. See the recipe in
# section 2.3 for details.
plot_shrinkage_comparison(fit$posterior, characteristic = "white") +
  ggplot2::labs(
    title    = "KRW race — 97 firms",
    subtitle = "Nonparametric vs linear shrinkage on theta_hat_race"
  )
```

### 7.2 Color override

```{r custom-color, fig.width = 7, fig.height = 5, fig.cap = "Decision frontier with a +theme tweak applied after the plot is built."}
# Lane-A protected figures: overriding fill/color forfeits the parity
# guarantee for that figure — see the Misconception note below.
# Safe customization on a Lane-A frontier: keep the protected aesthetics
# and adjust only the theme/layout layer.
plot_decision_frontier(
  observed        = fit$posterior,
  grid            = eb_posterior_grid(fit$estimates, fit$prior),
  classification  = fit$classification,
  characteristic  = "white",
  selection_share = 0.05
) +
  ggplot2::theme(legend.position = "bottom")
```

**Can I just use ggplot directly?**

> Yes — ebrecipe plots return ggplot objects so you can extend them.
> But the Lane-A protected family guarantees figure-level
> reproducibility against companion fixtures. Replace a geom on a
> Lane-A plot and you forfeit that guarantee.

## 8. Saving figures for publication

```{r save-demo}
# Writes a temporary PDF to verify the saving recipe end-to-end without
# polluting the working directory; readers should replace `out_path`
# with their preferred filename when adapting this for a paper.
p <- plot_decision_frontier(
  observed        = fit$posterior,
  grid            = eb_posterior_grid(fit$estimates, fit$prior),
  classification  = fit$classification,
  characteristic  = "white",
  selection_share = 0.05
) + theme_ebrecipe()

out_path <- file.path(tempdir(), "kr_w_frontier.pdf")
ggplot2::ggsave(out_path, p,
                width = 7, height = 5,
                device = cairo_pdf)  # cairo for UTF-8 fonts
file.exists(out_path)
```

## Practical pitfalls

1. **Composite vs single panels**. `plot_results()`, `plot_diagnostics()`,
   `plot_decision()` return patchwork compositions; `+ theme_*()`
   may behave unexpectedly. Apply `theme_ebrecipe()` to *individual*
   panels first when you need fine-grained control.
2. **Palette + manual scale**. Mixing `ebrecipe_palette()` with
   `scale_color_*()` follows ggplot's last-applied rule. Be explicit.
3. **PDF vs PNG**. UTF-8 box-drawing renders cleanly in PNG; for
   PDF, use `cairo_pdf` to avoid font-substitution warnings.

## Where to next

- **Workflow context**: `vignette("a2-discrimination-workflow")`
  shows these figures in their natural narrative position.
- **Replication & provenance**: `vignette("m5-replication-and-reproducibility")`
  documents the Lane-A protected target registry that backs the
  parity guarantee.

## Provenance

```{r provenance}
sessionInfo()
```

## References
