---
title: "M5: Replication and reproducibility - three contracts, audited"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  Reproducibility is a contract or it is nothing. This vignette runs
  the contract in front of you across three layers: (1) the frozen-core
  SHA256 contract for 12 source files (HF1) plus the 47-entry export
  ledger, (2) the `replication_mode = TRUE` 7-parameter lock that
  governs deterministic algorithm execution, and (3) the test fixture
  and protected-target registry that anchors numerical provenance.
  We deliberately separate the *what* (`eb_test()` is a threshold
  hypothesis-testing function, not a verification runner) from the
  *how* (`testthat` plus `inst/extdata/companion-parity/` carry the
  159 numerical targets). The vignette closes with a single-number
  provenance walk and a chain-of-custody appendix.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M5: Replication and reproducibility - three contracts, audited}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "ebrecipe"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. An auditor's eye — three provenance chains

This vignette is for the *auditor*. Three chains thread through the
package:

- **Code lineage**: Walters Stata + MATLAB → companion 06-01 R translation → ebrecipe v2 frozen-core SHA256
- **Number lineage**: MATLAB fixture → `inst/extdata/companion-parity/v1/registry/` → `testthat` assertion
- **Decision lineage**: 64 DEC catalog → app-c → `eb_control()` formals → vignette

```{r sessioninfo-1}
sessionInfo()
```

## 2. Contract 1 — Frozen-core SHA256 (HF1, 12 files) + 47-entry export ledger

The 12 frozen-core files implement the Walters parity path. Their
SHA256 hashes are locked in `inst/locked-core-checksums.txt`.

```{r hf1-locked-files}
locked_files_path <- system.file("LOCKED_FILES.md", package = "ebrecipe")
if (file.exists(locked_files_path) && nzchar(locked_files_path)) {
  hf1 <- readLines(locked_files_path)
  cat(head(hf1, 20), sep = "\n")
} else {
  cat("(LOCKED_FILES.md not packaged in this install)\n")
}
```

Verification (when run from package source root):

```{r hf1-sha-verify}
# Conceptual — the actual checksum manifest is in
#   inst/locked-core-checksums.txt
# and is verified by a pre-commit hook + testthat fixture.
manifest_path <- system.file("locked-core-checksums.txt", package = "ebrecipe")
if (file.exists(manifest_path) && nzchar(manifest_path)) {
  manifest <- readLines(manifest_path)
  cat("HF1 manifest has", length(manifest), "lines (12 file SHA256 each)\n")
} else {
  cat("(locked-core-checksums.txt not packaged in this install)\n")
}
```

The **47-entry export ledger** (`inst/EXPORTS_LEDGER.md`) is a parallel
locked artifact for the package's NAMESPACE.

## 3. Contract 2 — `replication_mode = TRUE` 7-parameter lock

When `replication_mode = TRUE`, `eb_control()` locks seven parameters
to the values that reproduce the Walters (2024) MATLAB output to
$10^{-3}$ tolerance.

```{r replication-mode-inspect}
ctrl <- eb_control(replication_mode = TRUE)
# Inspect the locked formals
str(ctrl, max.level = 1)
```

The seven locked parameters (from `R/eb.R:474-547` `.eb_replication_defaults()`):

| # | Parameter | Locked value | Role |
|---|-----------|--------------|------|
| 1 | `seed` | `1234L` | RNG initialization for stochastic guards |
| 2 | `n_knots` | `5L` | log-spline knot count |
| 3 | `n_grid` | `1000L` | support grid resolution |
| 4 | `mean_constraint` | `TRUE` | prior mean fixed to zero |
| 5 | `c_grid` | `seq(0.001, 0.15, by = 0.001)` (150 pts) | penalty search grid |
| 6 | `optimizer` | `"L-BFGS-B"` | numeric optimizer |
| 7 | `replication_mode` | `TRUE` | the switch itself |

Override attempt is caught by `.eb_warn_replication_override()`:

```{r override-warning, eval = FALSE}
# TODO: Phase 9 — illustrate the locked-formal override warning. The
# exact warning text varies by API revision; the canonical test lives
# in tests/testthat/test-eb_control-replication.R.
# eb_control(replication_mode = TRUE, n_knots = 7)
#> Warning: n_knots is locked under replication_mode = TRUE; reverting to 5.
```

## 4. Contract 3 — Test fixture + protected-target registry

The 159 numerical verification targets live in two places:

- **Lane A protected figures** (13 rows): `inst/extdata/companion-parity/v1/registry/protected-target-registry.csv`
- **Numerical targets**: `tests/testthat/` test files (run via `testthat::test_local()`)

```{r registry-inspect}
registry_path <- system.file(
  "extdata/companion-parity/v1/registry/protected-target-registry.csv",
  package = "ebrecipe"
)
if (file.exists(registry_path) && nzchar(registry_path)) {
  registry <- read.csv(registry_path, stringsAsFactors = FALSE)
  cat("Lane A protected targets:", nrow(registry), "rows\n")
  # Show whichever subset of typed columns are present
  keep_cols <- intersect(c("target_id", "view", "characteristic", "scale"),
                         names(registry))
  if (length(keep_cols)) print(head(registry[, keep_cols, drop = FALSE]))
} else {
  cat("(registry not packaged in this install)\n")
}
```

## 5. What `eb_test()` actually is

> **`eb_test()` 는 패키지를 검증하는 함수인가?**
>
> No. `eb_test()` is a *threshold hypothesis-testing* function for
> individual units — not a verification runner for the 159 targets.
> Its arguments are `threshold`, `alternative`, `fdr_level`, etc.
> Package-level verification runs via `testthat::test_local()` and the
> protected-target registry.

```{r eb-test-actual}
data(krw_firms)
# eb_test() example — hypothesis testing on individual units. Actual
# signature (verified): eb_test(x, s, threshold, alternative, fdr_level).
fit_test <- eb_test(
  x           = krw_firms$theta_hat_race,
  s           = krw_firms$se_race,
  threshold   = 0.05,
  alternative = "greater",
  fdr_level   = 0.05
)
# This tests whether each firm's true theta exceeds 0.05.
cat("Class       :", class(fit_test)[1], "\n",
    "Top-level slots:", paste(names(fit_test), collapse = ", "), "\n",
    "n units tested :", length(fit_test$theta_hat %||%
                                fit_test$posterior$.posterior_mean %||%
                                fit_test$pvalues %||% NA), "\n")
```

## 6. Stata/MATLAB → R: five documented translation deltas

The companion book (06-01 "Lessons Learned") documents five
cross-language deltas:

1. **File format**: Stata `.dta` → R `readstata13` for raw data
2. **Environment reproducibility**: `renv` lock vs MATLAB R201x
3. **Delta method porting**: hand-derived analytic forms re-encoded
4. **Bootstrap timing**: parallel B=2000 vs serial in MATLAB
5. **q-value convention**: raw Storey (R) vs monotonized (some MATLAB)

See `vignette("a4-diagnostics-and-standardization")` for the
implementation-side companion.

## 7. What chain-of-custody does NOT guarantee

Provenance is *recorded existence*, not *truth verification*. Four
lineage-break scenarios:

1. **Manifest 누락**: HF1 SHA passes but the manifest file itself
   isn't git-tracked
2. **Fixture rebuild without registry update**: `.rds` regenerated
   but `registry.csv` not synced
3. **Decision merge without DEC ID**: new design decision missing
   from `app-c` catalog
4. **R version drift**: `sessionInfo()` shows a different R version
   from the manifest record

Contract-first (SHA mismatch → commit reject) and provenance-first
(registry row missing → silent breakage) sound *different alarms*.

## 8. Provenance receipt appendix

```{r provenance-receipt}
sessionInfo()

# Single-number provenance walk on a hypothetical target
# (illustrative — actual receipt files live in registry path above)
cat("\n--- Provenance summary ---\n")
cat("HF1 frozen-core: 12 files\n")
cat("47-entry export ledger: locked\n")
cat("replication_mode: 7-parameter lock active when TRUE\n")
cat("Protected fixtures: 13 rows (Lane A)\n")
cat("Numerical targets: 159 in testthat\n")
```

## Where to next

- **Workflow application**: `vignette("a2-discrimination-workflow")`
  walks the recipe with `replication_mode = TRUE` to reproduce
  CD-78 (27 firms) bit-exactly.
- (m5 is terminal in the m-track spine — no forward link.)

## References

- Walters [-@walters2024empirical] — modern recipe + replication conventions
- Lee [-@lee2025walters_companion] — companion book lessons (06-01)
